---
title: "A tiered uncertainty analysis to assess uncertainty in a fact"
subtitle: "Work in progress RiskHunt3r"
author: "Ullrika Sahlin, Dawei Tang, ..."
format: 
  html:
    toc: true
    echo: false
    message: false
    warning: false
---

```{r}
#| echo: false
#| message: false

library(ggplot2)
library(SHELF)
```

```{r}
cm = "X"
```


## Overview

The purpose of an uncertainty analysis is to identify sources of uncertainty and evaluate their combined impact on the experts uncertainty in a categorical outcome of interest. 

This page describe an approach to quantify uncertainty examplified by hazard identification. 

A tiered uncertainty analysis can be

- an expert judgement on the outcome of interest directly

- an expert judgement on the outcome of interest considering uncertainties in streams of evidence (expressed qualitatively with rules how to combine them)

- an expert judgement on the outcome of interest considering uncertainties in streams of evidence (expressed quantitatively and a model how to combine them)

## Qualitative approaches 

Approaches resulting in a qualitative expression of uncertainty in hazard identification exists. These are also based on expert judgement, but where uncertainty is expressed qualitatively. 

A qualitative expression for uncertainty is more open for different interpretations compared to a quantitative expression on a well-defined scale. 

Qualitative expressions for uncertainty are helpful when making a judgement on the combined impact of sources to uncertainty. 

A qualitative expression of uncertainty in the hazard identification can be sufficient if this goes into a continued assessment, but if hazard identification is the only assessment or there is a conclusion to be made on the hazard identification before proceeding to the next step, then a quantitative expression of uncertainty is to recommend. 

## Qualitative -> Quantitative 

### Starting with 50% probability 



###  Starting with 0-100% probability 

Practical certainty 

## Fully quantitative 

Quality of evidence -> Quantitative likelihood function 

Proceed with prior + likelihood -> Posterior 

